A Computational Account of Default-Mode Function\\by Control Theory and Reinforcement Learning

### 2016-08-05 paper.tex
more

\subsection{Demystifying the free-energy principle and active-inference}
In this section we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in the form of the so-called  \textit{Helmholtz machine}. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For example the Friston theory axiomatizes that the brain uses a (problematic, as it implicitly assumes that posterior of each hidden unit is factorial, e.g) wake-sleep algorithm to train the underlying Helmholtz machine, etc.
\begin{table}[H]
  \begin{tabular}{p{2cm}|p{11cm}}
         \hline
         \textbf{variable}    & \textbf{short description}  \\ \hline
         $\langle E\rangle_p$ & Expectation (a.k.a average) of the
         random quantity $E$ w.r.t to the probability density $p$.\\ \hline
         $\mathcal H$ & The usual information-theoretic entropy operator.\\ \hline
         $D_{KL}(p||q)$ & The Kullback-Leibler divergence between the probability densities $p$ and $q$. \\ \hline
             $\zeta$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's internal state $b$ and sensory inputs $s$, i.e $\zeta = (s, b).$ \\ \hline
             $\psi$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
             $p_G(.|\zeta)$ & Generative density for ...\\ \hline
         $p_R(.|\zeta)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
         $F_G(\zeta)$ & Helmholtz free-energy for a model $p_G$ of generating the observation $\zeta$. This measures the surprise incured upon observing $\zeta$ generated by the model $G$.\\ \hline
         $F^R_G(\zeta)$ & Variational Helmholtz free-energy from model $R$
         to $G$.  Note that $F^G_G = F_G$.\\ \hline
  \end{tabular}
  \caption{Table of notations.}
\end{table}
\subsubsection{Helmholtz free-energy and the generative model}
Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
\begin{eqnarray}
  \begin{split}
    \text{generative surprise } &= -\log(p_G(\zeta)) = -\log(p_G(\zeta)) \times 1 = -\log(P_G(\zeta))\sum_{\psi}p_G(\psi |\zeta)\\
    &= -\sum_{\psi}p_G(\psi, \zeta)\log(p_G(\zeta))
    =-\sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi, \zeta)/p_G(\psi|\zeta))\\
    &= \sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi|\zeta))-\sum_{\psi}p_G(\psi |\zeta)\log  (p_G(\psi, \zeta))\\
    &= -\langle \log  (p_G(., \zeta)) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))\\
    &= \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta)) := F_G(\zeta)
  \end{split}
  \label{eq:helm}
\end{eqnarray}
where $E_G(\psi, \zeta)$ is the energy of an fictive thermodynamic system defined by setting
\begin{equation}
  E_G(\psi, \zeta) := -\log(p_G(\psi, \zeta)).
  \label{eq:gibbs}
\end{equation}
%% In the above, the variable $\psi$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
The last quantity $F_G(\zeta) := \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))$ in \eqref{eq:helm} is nothing other than the Helmholtz free-energy for the physical system correspond to the external world, prescribed by the generative density $p_G(\psi|\zeta)$ for the occupation of \textit{macrostates} $\psi$. Thus generative surprise and generative Helmholtz free-energy are different views on exactly the same object.

The goal of the agent is then to optimize over the generative model $G$ (i.e over the generative density $p_G(.|s)$, so as to minimize their surprise. It turns out that a direct attempt to attack this optimization problem by gradient descent on the free-energy $F_G(s)$ is furtile: the parameter update steps are not ``very clean'', and require rather cumbersome and heavy computations. A workaround is then to introduce a second density $p_R(.|\zeta)$ called a \textit{recognition} density to work in tandem with the generative density $p_G(.|\zeta)$: the former dreams / fantacizes whilst the latter tries to generate sensations which match these dreams! This primal-dual idea, first proposed in Hinton et al. 1995, is at the heart ot the general free-energy principle that we'll introduce shortly.

@@

Now, invoking \eqref{eq:gibbs} and applying Bayes rule, we get the Gibbs
distribution
\begin{equation}
  p_G(\psi|\zeta) = \frac{p_G(\psi|\zeta)}{p_G(\zeta)} = \frac{\exp(-E_G(\psi, \zeta))}{Z_G(\zeta)} = \frac{\exp(-E_G(\psi, \zeta))}{Z_G(\zeta)},
\end{equation}
where $Z_G(\zeta) := \log(p_G(\zeta)) = \sum_{\psi'}\exp(-E_G(\psi',\zeta))$, the normalizing partition function for the model \ref{eq:gibbs}.
Whence $\forall \psi$, $Z_G(\zeta) = p_G(\zeta) = \exp(-E_G(\psi, \zeta)) / p_G(\psi|\zeta)$, and so we have the invariance relation
\begin{equation}
  -\log(Z_G(\zeta)) = -\log(p_G(\zeta)) = F_G(\zeta) = E_G(\psi, \zeta) + \log(p_G(\psi|\zeta)).
  \end{equation}
Now, in the equation above, the LHS only depends on the generative model $G$ and the data point $\zeta$: it doesn't depend on the hidden variable $\psi$, etc. So, taking expectations w.r.t an arbitrary density\footnote{The conditioning in $P_R(.|\zeta)$ is because this density is selected from a world in which the sensory inputs and internal brain state vector $\zeta$ is assumed already observed.} $p_R(.|\zeta)$ yields
\begin{equation}
  \begin{split}
    F_G(\zeta) &= -\log(Z_G(\zeta)) = \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} + \sum_{\psi}p_R(\psi|\zeta)\log(p_G(\psi|\zeta))\\
    &= \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} - \mathcal H(p_R(.|\zeta)) - \sum_{\psi}p_R(\psi|\zeta)\log(p_R(\psi|\zeta)/p_G(\psi|\zeta))\\
    &= F^R_G(\zeta) - D_{KL}(P_R(.|\zeta) || P_G(.|\zeta)),
  \end{split}
  \label{eq:fe}
\end{equation}
where $F^R_G(\zeta)$ is the \textit{variational} Helmholtz free-energy from $R$ to $G$ defined by
\begin{equation}
  F^R_G(\zeta) := \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} - \mathcal H(p_R(.|\zeta))
\end{equation}
and $D_{KL}(P_R(.|\zeta) || P_G(.|\zeta))$ is the Kullback-Leibler divergence between the $p_R(.|\zeta)$ and the generative density $p_G(.|\zeta)$. Note that $F^G_G = F_G$.

\subsubsection{A general free-energy principle}
We can resume the situation as follows:
\begin{mdframed}
\begin{equation}
  \begin{split}
    \text{generative surprise } &:= -\log(p_G(\zeta)) = \underbrace{F_G(\zeta)}_{\text{generative free-energy}} \\
    &=\underbrace{F^R_G(\zeta)}_{\text{variational free-energy}} - \underbrace{D_{KL}(P_R(.|\zeta) || P_G(.|\zeta))}_{\text{complexity}} \\
    &\le F^R_G(\zeta),
    \text{ with equalitity if }p_R(.|\zeta) = p_G(.|\zeta),
    \end{split}
\end{equation}
\end{mdframed}
where we have used the fact that KL divergence is always nonnegative. In fact, we have the following theorem
\begin{theorem}
  It holds that
  $$
  \text{Minimal generative surprise = minimal (Helmholtz) variational free-energy},$$
  i.e
  \begin{equation}
    \min_{G}F_G(\zeta) = \min_{G,R}F^R_G(\zeta)
    \end{equation}
\end{theorem}

\subsubsection{Helmholtz machines}
...
\subsection{The wake-sleep algorithm}
...
\subsection{Friston's active-inference: an application of the Dayan's wake-sleep algorithm for training a Helmholtz machine model for the brain}
...

%% \subsection{A thermodynamic model for bounded rationality, aka robust optimality}
%% \begin{itemize}
%%   \item Recall that an agent is said to have \textit{bounded rationality} if they must take into account the cost of finding solutions to problems, and not just the utility of the final state.
%%   \item For example, consider an agent that must operate under a limited lifetime and/or computation cost.
%%   \item This is in contrast to agents with \textit{unbounded rationality} considered in classical game theory.
%%     \end{itemize}
%% The material presented is a revisit of \cite{braun2011path}. See also \cite{ortega2013thermodynamics}
%% \paragraph{Utility functions and conjugate pairs.}
%%   Let $(\Omega, \mathcal F, P)$ be a probablity space. A function $U: \mathcal F \rightarrow \mathbb R$ is said to be a \textit{utility function} for this space if the conditional utility $U(A|B) := U(A \cap B) - U(B)$ has the following propertites:
%%   \begin{itemize}
%%   \item additivity: $U(A_1 \cap A_2 | B) = U(A_1|B) + U(A_2|B)$, for all events $A_1, A_2, B \in \mathcal F$.    
%%   \item statistic: there exists a function $f_{U} :\mathbb R_+ \rightarrow \mathbb R$ such that $U(A|B) = f(P(A|B))$, for all events
%%     $A,B \in \mathcal F$.
%%   \item monotonicity: $f_{U}$ is strictly increasing.
%%     \end{itemize}

@@

\paragraph{Free Energy Principle.}
The free-energy principle in it's present form (including notions like ``generative density'', ``recognition density'', etc.) can be traceback to works of Dayan \& Hinton \cite{dayan1995helmholtz} in which the introduced the so-called Helmholtz machine...

\begin{itemize}
  \item Ortega et al. 2013 \cite{ortega2013thermodynamics} ``Thermodynamics as a theory of decision-making with information-processing costs''
\item Friston's critic of RL \cite{fristonAIorRL} (he proposes "active inference''): "This equation (equation 18) comes from the theory of dynamic programming,
pioneered in the 1950s by Richard Bellman and colleagues [2]. To
optimise control a~pð Þ x~ under this formulation, we have to: (i)
assume the hidden states are available to the agent and (ii) solve
Equation 18 for the value-function. Solving for the value-function
is a non-trivial problem and usually involves backwards induction
or some approximation scheme like reinforcement-learning [4–6].
The free-energy formulation circumvents these problems by
prescribing the policy in terms of free-energy, which encodes
optimal control (Equation 6)"
  \end{itemize}

>>>>

\subsection{Demystifying the free-energy principle and active-inference}
The free-energy principle in it's present form (including notions like ``generative density'', ``recognition density'', etc.) can be traceback to works of Dayan \& Hinton \cite{dayan1995helmholtz} in which the introduced the so-called Helmholtz machine...

In this section we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in the form of the so-called  \textit{Helmholtz machine}, a hierarchical factorial directional deep-belief net. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For example the Friston theory axiomatizes that the brain uses a (problematic, as it implicitly assumes that posterior of each hidden unit is factorial, e.g) wake-sleep algorithm to train the underlying Helmholtz machine, etc.
\begin{table}[H]
  \begin{tabular}{p{2cm}|p{11cm}}
         \hline
         \textbf{variable}    & \textbf{short description}  \\ \hline
         $\langle E\rangle_p$ & Expectation (a.k.a average) of the
         random quantity $E$ w.r.t to the probability density $p$.\\ \hline
         $\mathcal H$ & The usual information-theoretic entropy operator.\\ \hline
         $D_{KL}(p||q)$ & The Kullback-Leibler divergence between the probability densities $p$ and $q$. \\ \hline
             $\zeta$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's internal state $b$ and sensory inputs $s$, i.e $\zeta = (s, b).$ \\ \hline
             $\psi$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
             $p_G(.|\zeta)$ & Generative density for ...\\ \hline
         $p_R(.|\zeta)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
         $F_G(\zeta)$ & Helmholtz free-energy for a model $p_G$ of generating the observation $\zeta$. This measures the surprise incured upon observing $\zeta$ generated by the model $G$.\\ \hline
         $F^R_G(\zeta)$ & Variational Helmholtz free-energy from model $R$
         to $G$.  Note that $F^G_G = F_G$.\\ \hline
  \end{tabular}
  \caption{Table of notations.}
\end{table}
\subsubsection{Helmholtz free-energy and the generative model}
Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
\begin{eqnarray}
  \begin{split}
    \text{generative surprise } &= -\log(p_G(\zeta)) = -\log(p_G(\zeta)) \times 1 = -\log(P_G(\zeta))\sum_{\psi}p_G(\psi |\zeta)\\
    &= -\sum_{\psi}p_G(\psi, \zeta)\log(p_G(\zeta))
    =-\sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi, \zeta)/p_G(\psi|\zeta))\\
    &= \sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi|\zeta))-\sum_{\psi}p_G(\psi |\zeta)\log  (p_G(\psi, \zeta))\\
    &= -\langle \log  (p_G(., \zeta)) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))\\
    &= \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta)) := F_G(\zeta)
  \end{split}
  \label{eq:helm}
\end{eqnarray}
where $E_G(\psi, \zeta)$ is the energy of a fictive thermodynamic system defined by setting
\begin{equation}
  E_G(\psi, \zeta) := -\log(p_G(\psi, \zeta)).
  \label{eq:gibbs}
\end{equation}
%% In the above, the variable $\psi$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
The last quantity in \eqref{eq:helm} defined formally as
\begin{equation}
  F_G(\zeta) := \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))
\end{equation}
is the \textit{Helmholtz free-energy} of the external world, modelled as the physical system for which $p_G(\psi|\zeta)$ dictates the occupation probabilities of \textit{macrostates} $\psi$.
Thus generative surprise and generative Helmholtz free-energy are different views on exactly the same object.

The goal of the brain is then to optimize over the generative model $G$ (to iteratively or analytically modify the generative density $p_G(.|s)$, so as to minimize their surprise. It turns out that a direct attempt to attack this optimization problem by gradient descent on the free-energy $F_G(s)$ is furtile: the parameter update steps are not ``very clean'', and require rather cumbersome and heavy computations. A workaround is then to introduce a second density $p_R(.|\zeta)$ called a \textit{recognition} density to work in tandem with the generative density $p_G(.|\zeta)$: the former dreams / fantacizes whilst the latter tries to generate sensations which match these dreams! This primal-dual idea, first proposed in Hinton et al. 1995, is at the heart ot the general free-energy principle that we'll introduce shortly.

@@

Now, invoking \eqref{eq:gibbs} and applying Bayes rule, we get the Gibbs
distribution
\begin{equation}
  p_G(\psi|\zeta) = \frac{p_G(\psi|\zeta)}{p_G(\zeta)} = \frac{\exp(-E_G(\psi, \zeta))}{Z_G(\zeta)} = \frac{\exp(-E_G(\psi, \zeta))}{Z_G(\zeta)},
\end{equation}
where $Z_G(\zeta) := \log(p_G(\zeta)) = \sum_{\psi'}\exp(-E_G(\psi',\zeta))$, the normalizing partition function for the model \ref{eq:gibbs}.
Whence $\forall \psi$, $p_G(\zeta) = Z_G(\zeta) = \exp(-E_G(\psi, \zeta)) / p_G(\psi|\zeta)$, and so we have the invariance relation
\begin{equation}
F_G(\zeta)\overset{\eqref{eq:helm}}{=} -\log(p_G(\zeta)) = -\log(Z_G(\zeta)) = E_G(\psi, \zeta) + \log(p_G(\psi|\zeta)).
  \end{equation}
Now, in the equation above, the LHS only depends on the generative model $G$ and the data point $\zeta$: it doesn't depend on the hidden variable $\psi$, etc. So, taking expectations w.r.t an arbitrary density\footnote{The conditioning in $P_R(.|\zeta)$ is because this density is selected from a world in which the sensory inputs and internal brain state vector $\zeta$ is assumed already observed.} $p_R(.|\zeta)$ yields
\begin{equation}
  \begin{split}
    F_G(\zeta) &= -\log(Z_G(\zeta)) = \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} + \sum_{\psi}p_R(\psi|\zeta)\log(p_G(\psi|\zeta))\\
    &= \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} - \mathcal H(p_R(.|\zeta)) - \sum_{\psi}p_R(\psi|\zeta)\log(p_R(\psi|\zeta)/p_G(\psi|\zeta))\\
    &= F^R_G(\zeta) - D_{KL}(P_R(.|\zeta) || P_G(.|\zeta)),
  \end{split}
  \label{eq:fe}
\end{equation}
where $F^R_G(\zeta)$ is the \textit{variational} Helmholtz free-energy from $R$ to $G$ defined by
\begin{equation}
  F^R_G(\zeta) := \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} - \mathcal H(p_R(.|\zeta))
\end{equation}
and $D_{KL}(P_R(.|\zeta) || P_G(.|\zeta))$ is the Kullback-Leibler divergence between the $p_R(.|\zeta)$ and the generative density $p_G(.|\zeta)$. Note that $F^G_G = F_G$.

\subsubsection{A general free-energy principle}
We can resume the situation as follows:
\begin{mdframed}
\begin{equation}
  \begin{split}
    \text{generative surprise } &:= -\log(p_G(\zeta)) = \underbrace{F_G(\zeta)}_{\text{generative free-energy}} \\
    &=\underbrace{F^R_G(\zeta)}_{\text{variational free-energy}} - \underbrace{D_{KL}(P_R(.|\zeta) || P_G(.|\zeta))}_{\text{complexity}} \\
    &\le F^R_G(\zeta),
    \text{ with equalitity if }p_R(.|\zeta) = p_G(.|\zeta),
    \end{split}
\end{equation}
\end{mdframed}
where we have used the fact that KL divergence is always nonnegative.
%% In fact, we have the following theorem
%% \begin{theorem}
%%   It holds that
%%   $$
%%   \text{Minimal generative surprise = minimal (Helmholtz) variational free-energy},$$
%%   i.e
%%   \begin{equation}
%%     \min_{G}F_G(\zeta) = \min_{G,R}F^R_G(\zeta)
%%     \end{equation}
%% \end{theorem}

\subsubsection{Helmholtz machines}
% \begin{mdframed}
  \textbf{Assumption:} In both generative and recognition components of the network, there is conditional independence of  neurons in the same layer, given the data. Precisely
$$
p_G(\psi^{(l)}|\zeta) = \Pi_{k=1}^{h_l}p_G(\psi_k^{(l)} | \zeta),\hspace{1cm}
p_R(\psi^{(l)}|\zeta) = \Pi_{k=1}^{h_l}p_R(\psi_k^{(l)} |\zeta)
$$
% \end{mdframed}
\subsection{The wake-sleep algorithm}
...
\subsection{Friston's active-inference}
This is nothing but an application of the Dayan's wake-sleep algorithm for training a Helmholtz machine model of the brain...

The following critics can be made:
\begin{itemize}
  \item 
    As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
    \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
    \end{itemize}

\subsubsection{Free-energy minimization via a stochastic gradient variation bayes estimator}
\textbf{XXX see Kingma 2014 -- 2014: the poor boy is minimizing FE without knowing!}
\begin{equation}
  \begin{split}
    F_G^R(\zeta) &:= \langle E_G(., \zeta) \rangle_{p_R(. |\zeta)} - \mathcal H(p_R(. |\zeta)) = \langle E_G(., \zeta) - \log(p_R(.|\zeta))\rangle_{P_R(.|\zeta)}\\
    &=\langle \Delta E_{G \rightarrow R}(., \zeta)\rangle_{P_R(.|\zeta)}
    \end{split}
\end{equation}

For fixed $p_G(.|\zeta)$, the optimal choice for $p_R(.|\zeta)$ is given analytically by
\begin{equation}
  p_R(\psi|\zeta) \propto p_G(\psi|\zeta)\exp(-\Delta E_{G \rightarrow R}(\psi, \zeta)).
\end{equation}

%% \subsection{A thermodynamic model for bounded rationality, aka robust
%%   optimality}
%% \begin{itemize}
%%   \item Recall that an agent is said to have \textit{bounded rationality} if they must take into account the cost of finding solutions to problems, and not just the utility of the final state.
%%   \item For example, consider an agent that must operate under a limited lifetime and/or computation cost.
%%   \item This is in contrast to agents with \textit{unbounded rationality} considered in classical game theory.
%%     \end{itemize}
%% The material presented is a revisit of \cite{braun2011path}. See also \cite{ortega2013thermodynamics}
%% \paragraph{Utility functions and conjugate pairs.}
%%   Let $(\Omega, \mathcal F, P)$ be a probablity space. A function $U: \mathcal F \rightarrow \mathbb R$ is said to be a \textit{utility function} for this space if the conditional utility $U(A|B) := U(A \cap B) - U(B)$ has the following propertites:
%%   \begin{itemize}
%%   \item additivity: $U(A_1 \cap A_2 | B) = U(A_1|B) + U(A_2|B)$, for all events $A_1, A_2, B \in \mathcal F$.    
%%   \item statistic: there exists a function $f_{U} :\mathbb R_+ \rightarrow \mathbb R$ such that $U(A|B) = f(P(A|B))$, for all events
%%     $A,B \in \mathcal F$.
%%   \item monotonicity: $f_{U}$ is strictly increasing.
%%     \end{itemize}

@@

\paragraph{Free Energy Principle.}
\begin{itemize}
  \item Ortega et al. 2013 \cite{ortega2013thermodynamics} ``Thermodynamics as a theory of decision-making with information-processing costs''
\item Friston's critic of RL \cite{fristonAIorRL} (he proposes "active inference''): "This equation (equation 18) comes from the theory of dynamic programming,
pioneered in the 1950s by Richard Bellman and colleagues [2]. To
optimise control a~pð Þ x~ under this formulation, we have to: (i)
assume the hidden states are available to the agent and (ii) solve
Equation 18 for the value-function. Solving for the value-function
is a non-trivial problem and usually involves backwards induction
or some approximation scheme like reinforcement-learning [4–6].
The free-energy formulation circumvents these problems by
prescribing the policy in terms of free-energy, which encodes
optimal control (Equation 6)"
  \end{itemize}
